A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-14 text/abstract.Rmd
edit abstract

Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways relevant to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into ten classes distinct from their tissues-of-origin and largely
free of bias from mutational burden or clinical stage. We showed that these
tumor classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating that tissue-specific signatures contribute to common
cellular phenotypes. Moreover, driver gene mutations involved in replication fidelity and genome
stability, such as BRCA1 and Rb1, are equally distributed among our classes,
contrary to the expectation that such genes are avatars of molecular
subtype. We also investigated pathways that distinguish between metastatic and
non-metastatic tumors, and found that most meaningful differences are
class-specific. Many metastastasis enriched pathways from one class of
tumors are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on
for prevention of class-specific relapse. In summary, our
observations suggest that some genes act as drivers of genome instability in all
classes, while different sets of class-specific genes confer increased fitness
in a combinatorial fashion, eventually converging on one or more end-stage
disease phenotypes. Thus, we present a coherent view of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

>>>>

Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways relevant to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into ten classes distinct from their tissues-of-origin largely free of
bias from mutational burden or clinical stage. We showed that these tumor
classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating that tissue-specific signatures contribute to common
cellular phenotypes. Moreover, driver gene mutations involved in replication
fidelity and genome stability, such as BRCA1 and Rb1, are equally distributed
among our classes, contrary to the expectation that such genes are avatars of
molecular subtype. We also investigated pathways that distinguish between
metastatic and non-metastatic tumors, and found that most meaningful differences
are class-specific. Many metastasis enriched pathways from one class of tumors
are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on for
prevention of class-specific relapse. Taken as a whole, our observations suggest
that some genes act as drivers of the cancer phenotype in all classes, while
specific pathway disruptions confer increased fitness in a combinatorial
fashion, producing a limited number of distinct molecular class phenotypes.
Eventually these classes converge on one or more end-stage disease phenotypes.
Thus, we present a coherent view of global tumor biology, and explain how common
cellular dysfunction might arise from tissue-specific mutations.
